Three objections to a novel paradigm in social media effects research

### 2025-05-28 index.qmd
Add meta-psychology submission info

---
title: Three objections to a novel paradigm in social media effects research
running-head: Three objections
author:
  - name: Matti Vuorre
    orcid: 0000-0001-5052-066X
    email: mjvuorre@uvt.nl
    corresponding: true
    affiliation:
      - ref: 1
  - name: Niklas Johannes
    orcid: 0000-0001-6612-2842
    affiliation:
      - ref: 2
  - name: Andrew K. Przybylski
    orcid: 0000-0001-5547-2185
    affiliation:
      - ref: 2
affiliations:
  - id: 1
    name: Tilburg University
    department: Department of Social Psychology
  - id: 2
    name: University of Oxford
    department: Oxford Internet Institute
authornote: |
  This manuscript has been peer-reviewed and accepted for publication as a "research article" at [Scientific Reports](https://www.nature.com/srep/), but also subsequently rejected before peer-review as a "Matters Arising" article at the same journal. It is available under CC-BY 4.0 at <https://psyarxiv.com/dpuya/>.
abstract: |
  The study of social media effects on psychological well-being has reached an impasse: Popular commentators confidently assert that social media are bad for users but research results are mixed and have had little practical impact. In response, one research group has proposed a path forward for the field that moves beyond studying population averages to find effects that are specific to individuals. We outline three objections to that research agenda. On a methodological level, the key empirical results of this programme---proportions of the population of individuals with negative, null, and positive social media effects---are not appropriately estimated and reported. On a theoretical level, these results do little to advance our understanding of social media and its psychological implications. On a  paradigmatic level, this "personalized media effects paradigm" [@valkenburgSocialMediaUse2021, p.74] cannot inform inferences about individuals and therefore does not deliver what it claims. We express concern that this research approach may be contributing to confusing messaging to both societal stakeholders and scientists investigating how social media and well-being might be related. We sincerely hope that describing these objections prompts the field to work together in adopting better practices to develop a better understanding of well-being in the digital age.
keywords: [social media, media effects, well-being, variation, paradigm]
bibliography: person-specific-paradigm.bib
format:
  preprint-typst:
    output-file: ms.pdf
    wordcount: '{{< words-body >}}'
---

>>>>

---
title: Three objections to a novel paradigm in social media effects research
running-head: Three objections
author:
  - name: Matti Vuorre
    orcid: 0000-0001-5052-066X
    email: mjvuorre@uvt.nl
    corresponding: true
    affiliation:
      - ref: 1
  - name: Niklas Johannes
    orcid: 0000-0001-6612-2842
    affiliation:
      - ref: 2
  - name: Andrew K. Przybylski
    orcid: 0000-0001-5547-2185
    affiliation:
      - ref: 2
affiliations:
  - id: 1
    name: Tilburg University
    department: Department of Social Psychology
  - id: 2
    name: University of Oxford
    department: Oxford Internet Institute
authornote: |
  Submitted to Meta-Psychology. Participate in open peer review by sending an email to open.peer.reviewer@gmail.com. The full editorial process of all articles under review at Meta-Psychology can be found following this link: <https://tinyurl.com/mp-submissions>. You will find this preprint by searching for the first author's name.

  This manuscript has been peer-reviewed and accepted for publication as a "research article" at [Scientific Reports](https://www.nature.com/srep/), but also subsequently rejected before peer-review as a "Matters Arising" article at the same journal. It is available under CC-BY 4.0 at <https://psyarxiv.com/dpuya/>.
abstract: |
  The study of social media effects on psychological well-being has reached an impasse: Popular commentators confidently assert that social media are bad for users but research results are mixed and have had little practical impact. In response, one research group has proposed a path forward for the field that moves beyond studying population averages to find effects that are specific to individuals. We outline three objections to that research agenda. On a methodological level, the key empirical results of this programme---proportions of the population of individuals with negative, null, and positive social media effects---are not appropriately estimated and reported. On a theoretical level, these results do little to advance our understanding of social media and its psychological implications. On a  paradigmatic level, this "personalized media effects paradigm" [@valkenburgSocialMediaUse2021, p.74] cannot inform inferences about individuals and therefore does not deliver what it claims. We express concern that this research approach may be contributing to confusing messaging to both societal stakeholders and scientists investigating how social media and well-being might be related. We sincerely hope that describing these objections prompts the field to work together in adopting better practices to develop a better understanding of well-being in the digital age.
keywords: [social media, media effects, well-being, variation, paradigm]
bibliography: person-specific-paradigm.bib
format:
  preprint-typst:
    output-file: ms.pdf
    wordcount: '{{< words-body >}}'
---
